Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-05 paper/main.tex
Udpates

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remain limited, as researchers often prefer to construct DAGs manually
based on domain knowledge. This preference arises due to several practical
challenges with automated algorithms, such as their tendency to make obvious
errors and output Markov Equivalence Classes (MECs) rather than a DAG. To
address these challenges, we propose an iterative method that guides
researchers in manually constructing or improving the fit of an existing DAG.
Our approach combines implied Conditional Independence (CI) testing with a
novel measure of conditional association between variables to rank violations
of implied CIs in a given DAG. This ranking helps prioritize modifications to
the DAG to improve its fit to the data while utilizing expert knowledge to
decide the orientation of new edges. Empirical results show that this guided
manual model construction approach achieves performance comparable to automated
algorithms if we are able to correctly decide the orientation of new edges in
one out of three cases. Additionally, we demonstrate that, in the absence of an
expert, a Large Language Model (LLM) can be potentially used to determine edge
orientations. We provide the implementation of our approach in both a web tool
at: <redacted for review> and a Python package <redacted for review>.

@@

To tackle these issues with CI based testing, we take inspiration from
modification indices in specification search. The modification indices
method tries to identify and rank changes in the model such that modifications
that improve the model fit the most are ranked higher. This ranking allows
researchers to focus first on the suggestions that can improve the model the
most and while integrating their expertise to choose the most optimal
modifications. Analogous to this, we develop a procedure for ranking CI test
violations allowing researchers to focus on the worst violations in the model.
This ranking method utilizes both p-values and a measure of conditional
association. Utilizing this procedure researchers can iteratively construct
DAGs while integrating their domain knowledge in the model building procedure.

@@

Our main contributions in this paper are as follows:
\begin{enumerate}
	\item We propose a novel measure of conditional association for mixed
		data based on canonical correlations
		(Section~\ref{sec:mixed_association}. This measure of
		association generalizes some of the commonly used special case
		measures of association to mixed data.
	\item Using this conditional association measure, we present a
		procedure to manually construct DAGs by ranking violation of
		implied CIs in the model (Section~\ref{sec:modification}. This
		iterative procedure allows users to integrate their domain
		knowledge in the model construction procedure.
	\item To allow researchers to easily apply this method on their
		datasets, we present a web-browser based interactive tool as well
		as implementation in a Python package (Section~\ref{sec:web}).
	\item Lastly, we compare the manual DAG construction method with automated
		causal discovery algorithms (Section~\ref{sec:empirical}).

\end{enumerate}

@@

% As there are no generalized measures of association for mixed data, in this
% section, we first introduce a novel measure of partial association for mixed
% data by extending the concept of partial correlations \citep{Whittaker2009}
% \todo{Verify this citation} to handle mixed data types. To compute partial
% correlations, we have two key steps: 1) a residualization method, and 2) a
% measure of associations between the residuals. When both $ X $ and $ Y $ are
% continuous variables, residuals are typically computed using a regression-based
% method, and Pearson's correlation coefficient is used to measure the
% association between these residuals. To extend this approach to mixed data, we
% combine a mixed data residualization method \citep{Ankan2023} with a
% multivariate measure of association based on canonical correlations, Pillai's
% Trace \citep{Pillai1955}.

>>>>

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remain limited, as researchers often prefer to construct DAGs manually
based on domain knowledge. This preference arises due to several practical
challenges with automated algorithms, such as their tendency to make obvious
errors and output Markov Equivalence Classes (MECs) rather than a DAG. To
address these challenges, we propose an iterative method that guides
researchers in manually constructing or improving the fit of an existing DAG.
Our approach combines implied Conditional Independence (CI) testing with a
measure of conditional association between variables to rank violations
of implied CIs in a given DAG. This ranking helps prioritize modifications to
the DAG to improve its fit to the data while utilizing expert knowledge to
decide the orientation of new edges. Empirical results show that this guided
manual model construction approach achieves performance comparable to automated
algorithms if we are able to correctly decide the orientation of new edges in
one out of three cases. Additionally, we demonstrate that, in the absence of an
expert, a Large Language Model (LLM) can be potentially used to determine edge
orientations. We provide the implementation of our approach in both a web tool
at: <redacted for review> and a Python package <redacted for review>.

@@

To tackle these limitations, we draw inspiration from modification indices in
specification search. Modification indices provide ranking for potential model
modification based on their impact on the model's fit, allowing us to
prioritize most meaningful modifications. Similarly, we propose a method for
ranking CI test violations to help us focus on the most critical
inconsistencies in the model. To determine this ranking we utilize a measure of
conditional association between variables in combination with p-values from the
CI test. This approach helps us in tackling all the three issues outlined
above. The measure of association allows us to test for existing edges, as well
as not only rely on p-value to make decision whether a CI test holds. The
ranking helps us in focusing on the most critical violations.

@@

Our main contributions in this paper are as follows:

\begin{enumerate}
    \item We propose a novel measure of conditional association for mixed data
	    based on canonical correlations
	    (Section~\ref{sec:mixed_association}). This measure generalizes
	    several commonly used special case measures of association to mixed
	    data.
    \item Using this measure of conditional association, we develop a procedure
	    to rank violations of implied CIs. Using this ranking helps us
	    prioritize modifications to DAG to improve its fit while allowing
	    us to integrate expert knowledge when adding new edges (Section~\ref{sec:modification}).
    \item We provide a web-based interactive tool and a Python package to allow 
	    researchers to easily apply this method to their datasets (Section~\ref{sec:web}).
    \item Lastly, we compare the manual DAG construction method with automated
	    causal discovery algorithms (Section~\ref{sec:empirical}).
\end{enumerate}

@@
